
---
title: "Manual Refinement of Cell Type Annotations - Wu Dataset"
author: "Eric Lopez Huerta"
date: "2025-04-28"
output: html_document
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE)
```

# Introduction
This document refines and corroborates the cell type annotations of the Wu, 2021 dataset using SingleR and manual marker validation.

# Load Required Libraries
```{r}
library(Seurat)
library(ggplot2)
library(RColorBrewer)
library(SingleR)
library(dplyr)
library(cowplot)
library(gridExtra)
library(plyr)
```
# Load Input Data
```{r}
Annotated <- readRDS("data/objects/SnRAnnotWu_V4.rds")
References <- readRDS("data/objects/singleR_ReferenceMatrices.rds")
```

# SingleR Automatic Annotation
```{r}
Idents(Annotated) <- Annotated$seurat_clusters
Annotated$celltypes <- paste("SnR.", Annotated$celltypes, sep = "")
Annotated$Annotation <- Annotated$celltypes

reds <- brewer.pal(9, "YlOrRd")

Annotated <- readRDS("data/objects/IntegratedWu_V4.rds")
Annotated@meta.data[, c("celltype_subset", "celltype_minor", "celltype_major")] <- NULL

hESCs <- GetAssayData(Annotated, assay="RNA", layer='data')
bad_patterns <- 'RP[0-9]|^MT|[0-9][0-9][0-9][0-9]|^RPL|^RPS'
a <- rownames(hESCs)[!grepl(bad_patterns, rownames(hESCs))]
hESCs <- hESCs[a, ]

pred.hesc <- SingleR(test = hESCs, ref = References, method = "single", assay.type.test=1, labels = References$label.main)

# Adjust labels with low frequency
df.labels <- as.data.frame(table(pred.hesc$pruned.labels))
unknown <- as.character(df.labels[df.labels$Freq <= 100, 1])
new.labels <- ifelse(pred.hesc$pruned.labels %in% unknown, "unknown", pred.hesc$pruned.labels)
pred.hesc$labels <- new.labels

plotScoreHeatmap(pred.hesc)
saveRDS(pred.hesc, file = "results/singleR_results.rds")
```
# Manual Refinement of Annotations
```{r}
Clust.annot <- as.data.frame(cbind(rownames(pred.hesc), pred.hesc$labels))
new.cluster.ids <- Clust.annot$V2
names(new.cluster.ids) <- Clust.annot$V1
Annotated$celltypes <- new.cluster.ids
saveRDS(Annotated, file = "results/SnRAnnotWu_V4.rds")
```

# Identification of Marker Genes
```{r}
Cl.Mk <- FindAllMarkers(Annotated, assay = "RNA", slot = "data", logfc.threshold = 1, min.pct = 0.3, only.pos = TRUE)
top15 <- Cl.Mk %>% group_by(cluster) %>% filter(avg_log2FC > 2) %>% slice_head(n = 15) %>% ungroup()
```

# Median Expression Calculation
```{r}
S.Obj <- SplitObject(Annotated, split.by = "seurat_clusters")
med.list <- list()
for (i in seq_along(S.Obj)) {
  med.matrix <- GetAssayData(S.Obj[[i]], assay = "RNA", slot = "data")
  SubMark <- AllMarkers[AllMarkers %in% rownames(med.matrix)]
  med.list[[i]] <- apply(med.matrix[SubMark, ], 1, function(x) quantile(x, probs = 0.75))
}
names(med.list) <- names(S.Obj)
med.df <- plyr::ldply(med.list, rbind)
```

# Lymphoid Cell Refinement
```{r}
# (Lymphoid marker identification, plots, cluster refinements)
```

# Myeloid Cell Refinement
```{r}
# (Myeloid marker identification, plots, cluster refinements)
```

# Fibroblast, Epithelial, Endothelial Cells Refinement
```{r}
# (Fibroblast, epithelial and endothelial cluster refinements)
```

# Identification of Cycling Cells
```{r}
# (Detection and relabeling of cycling cells)
```

# Reassign Remaining Cells
```{r}
# (Assign rare or unassigned clusters based on majority identity)
```

# Save Final Annotated Seurat Object
```{r}
saveRDS(Annotated, file = "results/WuBRCA_Annotated.rds")
```

# Visualization
```{r}
# (DimPlots, FeaturePlots and DotPlots for final visualization)
```