
---
title: "TAMs Reclustering and Signature Analysis"
author: "Eric Lopez Huerta"
date: "2025-04-28"
output: html_document
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE)
```

# Introduction
This R Markdown document processes the Wu BRCA dataset by reclustering TAMs, identifying enriched signatures, and performing pathway analysis. Outputs include refined Seurat objects and plots.

# Load Libraries
```{r}
library(Seurat)
library(RColorBrewer)
library(ggplot2)
library(dplyr)
library(tibble)
library(stringr)
library(gridExtra)
library(org.Hs.eg.db)
library(limma)
library(cowplot)
library(plyr)
```

# Load Input Data
```{r}
paleta <- brewer.pal(9, "YlOrRd")
Annotated <- readRDS("data/objects/WuBRCA_Annotated.rds")
my_seurat <- readRDS("data/objects/Wu_Seurat_V4.rds")
```

# Interactive Selection
```{r}
Sub_Annot <- subset(Annotated, subset = celltypes %in% c("TAMs", "Cycling TAMs"))
TAMs.ID <- CellSelector(plot = DimPlot(Sub_Annot))
Doublets <- colnames(Sub_Annot)[!colnames(Sub_Annot) %in% TAMs.ID]
```

# First Reclustering
```{r}
my_seurat@meta.data[, c("celltype_subset", "celltype_minor", "celltype_major")] <- NULL
S.Obj <- my_seurat[, TAMs.ID]
rm(my_seurat)

samp.size <- sort(table(S.Obj$orig.ident), decreasing = TRUE)
samp.name <- names(samp.size)[samp.size <= 200]
first_half <- samp.name[1:(length(samp.name) / 2)]
second_half <- rev(samp.name)[1:(length(samp.name) / 2)]
fir_sec <- paste(first_half, second_half, sep = "_")

S.Obj$TAMsample <- as.character(S.Obj$orig.ident)
for (i in seq_along(first_half)) {
  S.Obj$TAMsample <- ifelse(S.Obj$TAMsample %in% c(first_half[i], second_half[i]), fir_sec[i], S.Obj$TAMsample)
}

Merg_list <- SplitObject(S.Obj, split.by = "TAMsample")
for (i in seq_along(Merg_list)) {
  Merg_list[[i]] <- NormalizeData(Merg_list[[i]], verbose = FALSE)
  Merg_list[[i]] <- FindVariableFeatures(Merg_list[[i]], selection.method = "vst", nfeatures = 2000, verbose = FALSE)
}

Merg.anchors <- FindIntegrationAnchors(object.list = Merg_list, dims = 1:30)
Merg.anchors <- IntegrateData(anchorset = Merg.anchors, dims = 1:30)
DefaultAssay(Merg.anchors) <- "integrated"
Merg.anchors <- ScaleData(Merg.anchors, verbose = FALSE)
Merg.anchors <- RunPCA(Merg.anchors, npcs = 30, verbose = FALSE)
Merg.anchors <- RunUMAP(Merg.anchors, reduction = "pca", dims = 1:30)
Merg.anchors <- FindNeighbors(Merg.anchors, dims = 1:10)
Merg.anchors <- FindClusters(Merg.anchors, resolution = 1)
```

# Refinement Based on Cluster Markers
```{r}
ClustMarkers <- FindAllMarkers(Merg.anchors, only.pos = TRUE, assay = "RNA", min.pct = 0.25, min.cells.feature = 15)
DF_Mark <- ClustMarkers %>%
  group_by(cluster) %>%
  slice_max(n = 20, order_by = avg_log2FC)

SIGNATURES <- lapply(split(DF_Mark, DF_Mark$cluster), function(df) head(df$gene, 10))
names(SIGNATURES) <- sapply(SIGNATURES, function(x) x[1])

Merg.anchors.scores <- AddModuleScore(Merg.anchors, features = SIGNATURES, ctrl = 20, name = "score", assay = "RNA")
Annotated_Score <- AddModuleScore(Annotated, features = SIGNATURES, ctrl = 19, name = "score", assay = "RNA")
```

# Pathway Enrichment Analysis
```{r}
PathDF <- getKEGGPathwayNames(species = "hsa", remove.qualifier = TRUE)
colnames(PathDF) <- c("ID", "Pathway")
```

# Second Reclustering
```{r}
BRCA_Refined <- SetIdent(Annotated, cells = Doublets, value = "Doublet")
BRCA_Refined$celltypes <- Idents(BRCA_Refined)
saveRDS(BRCA_Refined, file = "results/WuBRCA_Refined.rds")

my_seurat <- readRDS("data/objects/Wu_Seurat_V4.rds")
my_seurat@meta.data[, c("celltype_subset", "celltype_minor", "celltype_major")] <- NULL

TAMs.ID <- WhichCells(BRCA_Refined, idents = c("TAMs", "Cycling TAMs"))
TAMidents <- Idents(BRCA_Refined[, TAMs.ID])

S.Obj <- my_seurat[, TAMs.ID]
samp.size <- sort(table(S.Obj$orig.ident), decreasing = TRUE)
samp.name <- names(samp.size)[samp.size <= 233]
first_half <- samp.name[1:(length(samp.name) / 2)]
second_half <- rev(samp.name)[1:(length(samp.name) / 2)]
fir_sec <- paste(first_half, second_half, sep = "_")

S.Obj$TAMsample <- as.character(S.Obj$orig.ident)
for (ii in seq_along(first_half)) {
  S.Obj$TAMsample <- ifelse(S.Obj$TAMsample %in% c(first_half[ii], second_half[ii]), fir_sec[ii], S.Obj$TAMsample)
}

Merg_list <- SplitObject(S.Obj, split.by = "TAMsample")
for (i in seq_along(Merg_list)) {
  Merg_list[[ii]] <- NormalizeData(Merg_list[[ii]], verbose = FALSE)
  Merg_list[[ii]] <- FindVariableFeatures(Merg_list[[ii]], selection.method = "vst", nfeatures = 2000, verbose = FALSE)
}

Merg.anchors <- FindIntegrationAnchors(object.list = Merg_list, dims = 1:30)
Merg.anchors <- IntegrateData(anchorset = Merg.anchors, dims = 1:30)
DefaultAssay(Merg.anchors) <- "integrated"
Merg.anchors <- ScaleData(Merg.anchors, verbose = FALSE)
Merg.anchors <- RunPCA(Merg.anchors, npcs = 30, verbose = FALSE)
Merg.anchors <- RunUMAP(Merg.anchors, reduction = "pca", dims = 1:30)
Merg.anchors <- FindNeighbors(Merg.anchors, dims = 1:10)
Merg.anchors <- FindClusters(Merg.anchors, resolution = 0.5)

Merg.anchors <- SetIdent(Merg.anchors, cells = names(TAMidents), value = as.character(TAMidents))
Merg.anchors$celltypes <- Idents(Merg.anchors)
saveRDS(Merg.anchors, file = "results/WuTAMs.rds")
```
